---
title: "SE3 – Statistical Modeling of Emerging Contaminants"
author: "Robel Bekele"
date: "`r Sys.Date()`"
output: html_document
---

## Overview

This section presents statistical modeling of emerging contaminant concentrations in wastewater treatment plant (WWTP) effluents. The analysis uses processed data generated in SE2 and applies linear models to evaluate the influence of season, treatment technology, and WWTP variability.

## Load libraries

```{r}
library(readr)
library(dplyr)
library(broom)
```

## Load Processed Data


```{r}
MPs_WWTPs_Eff_DUMMY = read_csv("~/data/processed/analysis_ready_MPs_data.csv")

# Clean column names
colnames(MPs_WWTPs_Eff_DUMMY) <- make.names(colnames(MPs_WWTPs_Eff_DUMMY))

# Preview data
head(MPs_WWTPs_Eff_DUMMY)
```

# Data preparation

```{r}
# Select contaminant columns
cont_cols <- colnames(MPs_WWTPs_Eff_DUMMY)[11:(ncol(MPs_WWTPs_Eff_DUMMY)-1)]

# Log transformation
MPs_WWTPs_Eff_DUMMY_trans <- MPs_WWTPs_Eff_DUMMY

for (col in cont_cols) {
  MPs_WWTPs_Eff_DUMMY_trans[[col]] <- log10(MPs_WWTPs_Eff_DUMMY[[col]] + 1)
}

# Convert variables to factors
MPs_WWTPs_Eff_DUMMY_trans$Season <- factor(MPs_WWTPs_Eff_DUMMY_trans$Season)
MPs_WWTPs_Eff_DUMMY_trans$WWTP <- factor(MPs_WWTPs_Eff_DUMMY_trans$WWTP)
MPs_WWTPs_Eff_DUMMY_trans$Tech_type <- factor(MPs_WWTPs_Eff_DUMMY_trans$Tech_type)
```

# Model Training

```{r}
models <- list()

for (col in cont_cols) {
  
  formula <- as.formula(
    paste0("`", col, "` ~ Season + Tech_type + WWTP")
  )
  
  models[[col]] <- tryCatch(
    lm(formula, data = MPs_WWTPs_Eff_DUMMY_trans),
    error = function(e) NULL
  )
}
```

# Extract model outputs

```{r}
summary_table <- data.frame()

for (col in cont_cols) {
  
  if (!is.null(models[[col]])) {
    
    tidy_out <- tidy(models[[col]])
    
    season_p <- min(
      tidy_out$p.value[grepl("^Season", tidy_out$term)],
      na.rm = TRUE
    )
    
    tech_p <- min(
      tidy_out$p.value[grepl("^Tech_type", tidy_out$term)],
      na.rm = TRUE
    )
    
    summary_table <- rbind(
      summary_table,
      data.frame(
        Contaminant = col,
        Season_p_value = season_p,
        Technology_p_value = tech_p,
        Season_Significant = ifelse(season_p < 0.05, "Yes", "No"),
        Tech_Significant = ifelse(tech_p < 0.05, "Yes", "No")
      )
    )
  }
}

head(summary_table)
```

# Model Fit metrices

```{r}
model_fit <- data.frame()

for (col in cont_cols) {
  
  if (!is.null(models[[col]])) {
    
    g <- glance(models[[col]])
    
    model_fit <- rbind(
      model_fit,
      data.frame(
        Contaminant = col,
        R_squared = g$r.squared,
        Adj_R_squared = g$adj.r.squared,
        AIC = g$AIC
      )
    )
  }
}

head(model_fit)
```
