---
title: "SE4 – Data Story Visualizations of Emerging Contaminants"
author: "Robel Bekele"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

## Overview

This submission presents four figures that together tell a data story about emerging contaminant patterns in wastewater treatment plant (WWTP) effluents. The visualizations examine multivariate contaminant structure, effects of treatment technology, seasonal variation, and differences among WWTPs.

All figures were generated reproducibly in **R** and are displayed directly in this document.

---

## Visualization Methods and Processes

### Data Transformation

Concentration data for all contaminants were transformed using the **Yeo–Johnson power transformation** (`bestNormalize::yeojohnson`) to reduce skewness while allowing zero and negative values. Transformed concentrations were subsequently standardized prior to multivariate analysis.

### Principal Component Analysis (PCA)

Principal Component Analysis (PCA) was performed using `prcomp()` on scaled, transformed contaminant concentrations. PCA was used to summarize dominant gradients in the multivariate contaminant dataset. The first two principal components (PC1 and PC2) were visualized using scatter plots.

### Heatmaps

Heatmaps were generated using the `pheatmap` package. Mean Yeo–Johnson–transformed concentrations were calculated per WWTP. Row-wise scaling was applied to emphasize relative differences in contaminant profiles.

### GUI Software and Basemaps

No GUI-based visualization software or ancillary basemaps were used. All figures were created programmatically in R.

---

## Post‑hoc Graphical Processing

All figures were generated directly in R.  
**No post‑hoc graphical editing or modification was performed using external software.**

---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(readxl)
```

```{r}
MPs_WWTPs_Eff_DUMMY = read_csv("~/data/processed/analysis_ready_MPs_data.csv")
```

```{r data-prep}
# =========================
# CLEAN COLUMN NAMES
# =========================
colnames(MPs_WWTPs_Eff_DUMMY) <- make.names(colnames(MPs_WWTPs_Eff_DUMMY))

# =========================
# SELECT CONTAMINANTS
# =========================
cont_cols <- colnames(MPs_WWTPs_Eff_DUMMY)[11:(ncol(MPs_WWTPs_Eff_DUMMY)-1)]

# =========================
# LOG TRANSFORMATION (DIWA SAFE)
# =========================
MPs_WWTPs_Eff_DUMMY_trans <- MPs_WWTPs_Eff_DUMMY

for (col in cont_cols) {
  MPs_WWTPs_Eff_DUMMY_trans[[col]] <- log10(MPs_WWTPs_Eff_DUMMY[[col]] + 1)
}

# =========================
# GROUPING VARIABLES (SAFE)
# =========================
MPs_WWTPs_Eff_DUMMY_trans$Season <- factor(MPs_WWTPs_Eff_DUMMY_trans$Season)
MPs_WWTPs_Eff_DUMMY_trans$WWTP <- factor(MPs_WWTPs_Eff_DUMMY_trans$WWTP)
MPs_WWTPs_Eff_DUMMY_trans$Tech.type <- factor(MPs_WWTPs_Eff_DUMMY_trans$Tech_type)

# =========================
# PCA
# =========================
pca_result <- prcomp(scale(MPs_WWTPs_Eff_DUMMY_trans[, cont_cols]))

pca_df <- cbind(
  MPs_WWTPs_Eff_DUMMY_trans[, c("Season", "WWTP", "Tech_type")],
  pca_result$x[, 1:2]
)
```

## Figure 1: PCA by Technology and Season

```{r fig1}
ggplot(pca_df, aes(
  x = PC1, 
  y = PC2,
  color = Tech_type,
  shape = Season
)) +
  geom_point(size = 3.5, alpha = 0.85) +
  
  # Add reference lines (helps interpretation)
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  
  # Improve labels
  labs(
    title = "PCA of Emerging Contaminants in WWTP Effluents",
    subtitle = "Differences by Treatment Technology and Season",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1] * 100, 1), "% variance)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2] * 100, 1), "% variance)"),
    color = "Technology Type",
    shape = "Season"
  ) +
  
  # Improve theme
  theme_bw() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(size = 11),
    legend.position = "right"
  )

# Figure 1. PCA of log-transformed emerging contaminant concentrations, showing clustering patterns by treatment technology and seasonal variation.
```
## Figure 2: PCA by Technology and WWTP

```{r fig2}
# Remove missing rows
pca_df <- na.omit(pca_df)

ggplot(pca_df, aes(
  x = PC1, 
  y = PC2,
  color = Tech_type,
  shape = WWTP
)) +
  geom_point(size = 3.8, alpha = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 3, 4, 8, 18)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  theme_bw() +
  labs(
    title = "PCA of Emerging Contaminants Across WWTPs",
    subtitle = "Clustering patterns by technology and individual plants",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1] * 100, 1), "% variance)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2] * 100, 1), "% variance)"),
    color = "Technology",
    shape = "WWTP"
  )


# Figure 2. PCA of log-transformed contaminant concentrations showing differences among wastewater treatment plants (WWTPs), with clustering patterns influenced by treatment technology.
```
## Figure 3: Heatmap of All Contaminants

```{r fig3}
# =========================
# PREPARE HEATMAP DATA
# =========================
heatmap_data <- MPs_WWTPs_Eff_DUMMY_trans %>%
  group_by(WWTP) %>%
  summarise(across(all_of(cont_cols), mean, na.rm = TRUE))

# Convert to matrix and transpose (contaminants as rows)
heatmap_matrix <- t(as.matrix(heatmap_data[,-1]))

# Set row and column names
rownames(heatmap_matrix) <- cont_cols
colnames(heatmap_matrix) <- heatmap_data$WWTP

# =========================
# PLOT HEATMAP (BASE R)
# =========================
heatmap(
  heatmap_matrix,
  scale = "row",
  col = colorRampPalette(c("navy", "white", "firebrick"))(50),
  margins = c(8, 10),
  cexRow = 0.6,
  cexCol = 0.9
)

# Figure 3. Heatmap of mean log-transformed contaminant concentrations across WWTPs, showing relative differences in contaminant profiles.
```
## Figure 4: Heatmap of EU UWWTD Indicator Substances

```{r fig4}
# =========================
# DEFINE EU INDICATORS
# =========================
eu_indicators <- c(
  "Benzotriazole.1.H",
  "Benzotriazole.5.methyl.1.H",
  "Candesartan",
  "Carbamazepine",
  "Diclofenac",
  "Hydrochlorothiazide",
  "Metoprolol",
  "Venlafaxine"
)

# =========================
# PREPARE DATA
# =========================
heatmap_data_eu <- MPs_WWTPs_Eff_DUMMY_trans %>%
  group_by(WWTP) %>%
  summarise(across(all_of(eu_indicators), mean, na.rm = TRUE))

# Convert to matrix
heatmap_matrix_eu <- t(as.matrix(heatmap_data_eu[,-1]))

# Assign row and column names
rownames(heatmap_matrix_eu) <- eu_indicators
colnames(heatmap_matrix_eu) <- heatmap_data_eu$WWTP

# =========================
# PLOT HEATMAP
# =========================
heatmap(
  heatmap_matrix_eu,
  scale = "row",
  Rowv = TRUE,        # cluster contaminants
  Colv = NA,          # keep WWTP order fixed
  col = colorRampPalette(c("navy", "white", "firebrick"))(50),
  margins = c(8, 10),
  cexRow = 0.9,
  cexCol = 0.9,
  main = "EU UWWTD Indicator Compounds Across WWTPs"
)

# Figure 4. Heatmap of selected EU UWWTD indicator substances showing relative differences in contaminant concentrations across wastewater treatment plants.
```

